---
title: "ECTIL: prognostic PARADIGM analysis"
author: "Yoni Schrirris, Renee Menezes"
date: "`r Sys.Date()`"
params:
  dataset_file: "/path/to/data.RData"
  prediction_column: "ectil_label_column"
output:
    html_document:
    code_folding: hide
toc: true
toc_float: true
toc_collapsed: true
toc_depth: 3  
---

This is part of the prognostic analysis of "ECTIL: Label-efficient Computational Tumour Infiltrating Lymphocyte (TIL) assessment in breast cancer: Multicentre validation in 2,340 patients with breast cancer"

This file performs the survival analysis on PARADIGM using any of the 
predicted scores, and compares it to the pathologist's scores.

Loading the required packages
```{r}
library(ggplot2)
if (!require(cmprsk)) {
  install.packages("cmprsk")
}
if (!require("xtable")){
    install.packages("xtable")
}

# For tidying the tables and making them easily copy pastable.
if(!require("tidyr")){
    install.packages("tidyr")
}

# Same, also for cox reg. tables
if(!require("dplyr")){
    install.packages("dplyr")
}

# Same, also for cox reg. tables
if(!require("ggsurvfit")){
    install.packages("ggsurvfit")
}
library(ggsurvfit)

if (!require("survminer")) {
    install.packages("survminer")
}

library("survminer")
library("dplyr")
library("xtable")
library(cmprsk) # Competing risk analyses
library(gplots) # nicer plotting
library(readxl) # Reading excels
library(survminer)
library(ggplot2)
library(gridExtra) # tableGrob
library(survival) # necessary for all survival analyses
library(broom) # Nice for repackaging the survival results into a table
if (!require("data.table")) {
    install.packages("data.table")
}
library(data.table) # Nice for ":=" operator when reorganizing our tables for the cox regressions 
```

Load the dataset with all scores. This is then found in `data.all.scores`.
These can not be publicly shared, hence the analysis can only be reproduced within the Netherlands Cancer Institute.
```{r}
load(params$dataset_file)
```

We reproduce the analysis from 
de Jong VMT, Wang Y, ter Hoeve ND, et al. Prognostic Value of Stromal 
Tumor-Infiltrating Lymphocytes in Young, Node-Negative, Triple-Negative Breast 
Cancer Patients Who Did Not Receive (neo)Adjuvant Systemic Therapy. JCO 2022; 
JCO.21.01536.

Results may differ slightly since we only use the patients that we have WSIs for and were able
to predict a score for, hence we have fewer patients.

Definition of time and event variables
```{r}
# Define a time variable
data.all.scores$time_variable = as.numeric(as.Date(data.all.scores$date.last.follow.up, format="%d/%m/%Y") - as.Date(data.all.scores$date_diagnosis, format="%d/%m/%Y"))

# Define the event variable
data.all.scores$event_variable = data.all.scores$surv.stat

# For OS we will treat 2 (lost to follow-up) as censored
data.all.scores$event_variable[data.all.scores$event_variable==2] = 0

# Define a response variable
data.all.scores$response_variable = Surv(time=data.all.scores$time_variable, event=data.all.scores$event_variable)

# From days to years
data.all.scores$time_variable = data.all.scores$time_variable / 365
```

Set up a function to fit the multivariate coxph
```{r}
fitCoxPHModel <- function(data, variable_of_interest, variates) {
  coxph(as.formula(paste("response_variable ~", variable_of_interest, variates
                         )
                   ), 
        data = data)
}
```

All cox regressions; univariates with pathologist and model predictions,
multivare with no tils
multivariate with pathologist and mdoel predictions
```{r}

multivariate = "+
                trinarized_tumor_stage +
                binarized_grade +
                trinarized_histology +
                  `Lymphovascular.invasion` + 
                  radiotherapy_surgery"


univariate_fit_pathologist_stils <- fitCoxPHModel(data.all.scores, "increment_stils_10", "")
univariate_fit_weakstil_small <- fitCoxPHModel(data.all.scores, "increment_weakstils_small_10", "")
univariate_fit_weakstil_large_all <- fitCoxPHModel(data.all.scores, "increment_weakstils_large_all_10", "")
univariate_fit_weakstil_large_tnbc <- fitCoxPHModel(data.all.scores, "increment_weakstils_large_tnbc_10", "")


multivariate_fit_no_stils <- fitCoxPHModel(data.all.scores, "", multivariate)
multivariate_fit_pathologist_stils <- fitCoxPHModel(data.all.scores, "increment_stils_10", multivariate)
multivariate_fit_weakstil_small <- fitCoxPHModel(data.all.scores, "increment_weakstils_small_10", multivariate)
multivariate_fit_weakstil_large_all <- fitCoxPHModel(data.all.scores, "increment_weakstils_large_all_10", multivariate)
multivariate_fit_weakstil_large_tnbc <- fitCoxPHModel(data.all.scores, "increment_weakstils_large_tnbc_10", multivariate)
```

All models are fit now, but not shown.

We process them to format them in paper format.

```{r}
# Assuming 'fit' is your CoxPH model

processCoxPHModel <- function(fit, column_name, add_to = NULL) {
    conf_int <- summary(fit)$conf.int
    tidy_fit <- tidy(fit) # Comes from the broom package. Nicely represents this stuff in df.
    # Add confidence intervals and formatted output
    tidy_fit <- tidy_fit %>%
    mutate(exp_coef = exp(estimate),  # Exponentiated estimates
          se = std.error,
          lower_ci = conf_int[, "lower .95"],
          upper_ci = conf_int[, "upper .95"],
          p_signif = case_when(
            p.value < .001 ~ "p<0.001",
            p.value < .01 ~ "p<0.01",
            p.value < .05 ~ "p<0.05",
            TRUE ~ ""
          ),
          exp_coef = format(round(exp_coef, 2), nsmall = 2),  # Round to 2 decimals
          lower_ci = format(round(lower_ci, 2), nsmall = 2),
          upper_ci = format(round(upper_ci, 2), nsmall = 2),
          exp_coef_p = ifelse(p_signif == "", 
                              paste0(exp_coef, " (", lower_ci, " to ", upper_ci, ")"), 
                              paste0(exp_coef, " (", lower_ci, " to ", upper_ci, ", ", p_signif, ")"))
    )
    # Calculate the Concordance Index (C-index)
    c_index <- concordance(fit)$concordance
    c_index_se <- sqrt(concordance(fit)$var)
    
    # Compute the significance of it differing from 0.5
    null_value = 0.5
    Z = (c_index - null_value) / c_index_se
    p_value = (1-pnorm(abs(Z))) # One-sided
    
    p_signif = case_when(
           p_value < .001 ~ " (p<0.001)", # set significane levels
           p_value < .01 ~ " (p<0.01)",
           p_value < .05 ~ " (p<0.05)",
           TRUE ~ ""  # No asterisk for p >= 0.05
    )
    
    c_index_formatted <- paste0(sprintf("%.2f", c_index), p_signif)
    
    c_index_row <- tibble(variate = "Concordance Index", !!column_name := c_index_formatted)

    tidy_fit <- tidy_fit %>%
        select(variate = term,  !!column_name := exp_coef_p) # Term is how it is called normally. We call it "variate".

    extra_row <- tibble(variate = "Concordance Index", !!column_name := as.character(c_index_formatted))
    
    tidy_fit <- bind_rows(tidy_fit, extra_row)

    if (!is.null(add_to)){
        add_to[[column_name]] = tidy_fit[[column_name]]
        return(add_to)
    } else{
        return(tidy_fit)
    }
}

# We produce a nice table, and add every extra model to it to have a complete
# cox regression overview.
tidy_fit = processCoxPHModel(multivariate_fit_no_stils, "Clinical")

tidy_fit # Show it in Rmd file

tidy_fit = processCoxPHModel(multivariate_fit_pathologist_stils, "Pathologist sTIL")
tidy_fit = processCoxPHModel(multivariate_fit_weakstil_small, "ECTIL-TCGA", add_to=tidy_fit)
tidy_fit = processCoxPHModel(multivariate_fit_weakstil_large_all, "ECTIL-all", add_to=tidy_fit)
tidy_fit = processCoxPHModel(multivariate_fit_weakstil_large_tnbc, "ECTIL-TNBC", add_to=tidy_fit)


tidy_fit # Show it in Rmd file

tidy_fit = processCoxPHModel(univariate_fit_pathologist_stils, "sTIL-gt")
tidy_fit = processCoxPHModel(univariate_fit_weakstil_small, "ECTIL-TCGA", add_to=tidy_fit)
tidy_fit = processCoxPHModel(univariate_fit_weakstil_large_all, "ECTIL-all", add_to=tidy_fit)
tidy_fit = processCoxPHModel(univariate_fit_weakstil_large_tnbc, "ECTIL-TNBC", add_to=tidy_fit)

tidy_fit # Show it in Rmd file
```


Plot all KM plots
```{r}
# Assuming you have already installed and loaded the survminer and ggplot2 packages
# Plot survival curves for each level of stil_category as defined before
# Adjusting the Kaplan-Meier plot for better aesthetics

plot_custom_km = function(fit, group_labels, group_palette, filename){
    # Generate Kaplan-Meier plot with basic customization
    km_plot <- ggsurvplot(
      fit, 
      risk.table = "nrisk_cumcensor", # Show risk table
      tables.y.text = FALSE, 
      tables.theme = theme_cleantable(),
      pval = FALSE,  # Do not show the significance
      pval.coord = c(0,0), # p-value location
      conf.int = FALSE, # Do not distract too much with the confidence interval
      palette = group_palette, # Set the colours as given by user
      xlab = "Time (Years)", 
      ylab = "OS",
      xlim = c(0, 15), # Show for 15 years
      break.time.by = 1, # Show tick every 1 year
      ggtheme = theme_minimal() + # Set styling
        theme(
          plot.title = element_text(size = 14, face = "bold", color = "black"),
          axis.title.x = element_text(size = 12, face = "bold", color = "black"),
          axis.title.y = element_text(size = 12, face = "bold", color = "black"),
          axis.text.x = element_text(size = 10, face = "plain", color = "black"),
          axis.text.y = element_text(size = 10, face = "plain", color = "black"),
          panel.grid.major = element_blank(), #disable grid
          panel.grid.minor = element_blank(), # disable grid
          axis.line = element_line(color = "black"),
        ),
      risk.table.height = 0.2, # Make the table readable
      size = 0.6, # Adjust line thickness
      risk.table.fontsize = 3, # Adjust font size for the risk table
      legend.labs = group_labels, # Customize legend labels if needed,
      legend = c(0.85,0.2),
      legend.title=""
    )
    
    # Make the risk table super minimal
    km_plot$table <- km_plot$table + theme(
        plot.title = element_text(size = 8, face="plain"),
        )

    km_plot$table <- km_plot$table + labs(title = "No. at risk (censored)")
    
    ## Add the table with CIs
    
    group_names <- names(fit$strata)

    # Extract and format survival probabilities for each time point
    get_surv_at_times <- function(fit_obj, time_points) {
      # Initialize a list to hold survival data for each group
      surv_data <- list()
      surv_info <- summary(fit_obj, times = time_points, censored = TRUE)
      # Loop over each group in the fit object
      for (j in 1:length(group_names)) {
        # Format the survival, lower, and upper CI data
        formatted_surv_info <- sapply(1:length(time_points), function(i) {
          idx = (j-1)*3+i # Get specific items for the table, and format them
          sprintf("%.1f (%.1f to %.1f)", surv_info$surv[idx] * 100, surv_info$lower[idx] * 100, surv_info$upper[idx] * 100)
        })
        
        # Add the formatted survival information to the list
        group_name=group_names[j]
        surv_data[[group_name]] <- formatted_surv_info
      }
      return(surv_data)
    }
    
    # Specify time points (years converted to days)
    time_points <- c(3, 10, 15)
    
    # Get survival data for each group at specified time points
    survival_data <- get_surv_at_times(fit, time_points)
    
    # Create a dataframe for the custom table
    custom_table <- data.frame(
      Group = group_names,
      `3 Years` = sapply(group_names, function(g) survival_data[[g]][1]),
      `10 Years` = sapply(group_names, function(g) survival_data[[g]][2]),
      `15 Years` = sapply(group_names, function(g) survival_data[[g]][3])
    )
    
    names(custom_table) = c("TILs level", "3 Years (95% CI)", "10 Years (95% CI)", "15 Years (95% CI)")
    custom_table$`TILs level` = group_labels
    
    
    
    # Starting base theme
    table_theme <- ttheme_default(
      core = list(
        fg_params = list(hjust = 0, x = 0.1, col="black"), # Foreground parameters for text alignment
        bg_params = list(fill = "white", col = "black") # Background fill and border color
      ),
      colhead = list(
        fg_params = list(hjust = 0, x = 0.1), # Column header text alignment
        bg_params = list(fill = "white", col = "black") # Column header background and border
      ),
      base_size = 6, # Base font size for text in the table
      padding = unit(c(4, 4), "mm"), # Padding inside cells
      spacing = unit(c(2, 2), "mm") # Spacing between cells
    )
    colors_for_a <- group_palette # Example colors for each cell of 1st column

    # Adjust the theme to apply these colors
    white_colors = c("white", "white", "white")
    table_theme$core$bg_params$fill <- rep("white", ncol(custom_table) * nrow(custom_table))
    table_theme$core$bg_params$fill[seq(1, by = 1, length.out = nrow(custom_table))] <- colors_for_a
    table_theme$core$fg_params$col[seq(1, by = 1, length.out = nrow(custom_table))] <-  rep("white", length(colors_for_a))
    table_theme$core$fg_params$col[seq(nrow(custom_table)+1, by = 1, length.out = nrow(custom_table)*ncol(custom_table))] <-  rep("black", nrow(custom_table) * ncol(custom_table) - nrow(custom_table))
    table_theme$core$bg_params$col <- "black" # Ensure border colors are set to black
        
    # Create the table grob
    table_grob <- tableGrob(custom_table, theme = table_theme, rows=NULL)
    
    # Add the table grob to the KM plot, adjusting position
    km_plot$plot <- km_plot$plot +
      annotation_custom(
        grob = table_grob,
        xmin = unit(5, "npc"), ymin = unit(0.11, "npc"), xmax = unit(5.2, "npc"), ymax = unit(0.26, "npc")
      )
    # Set width, height DPI. If this is changed, will likely have to adjust
    # The line thickness and font size and so on.
    dpi=300
    width= 600 * dpi/72
    height = 500 * dpi/72
    
    print(km_plot)
    dev.copy(png, filename = filename, width = width, height = height, res=dpi)
    dev.off()
}
```

Standard KM plot with pathologist scores with normal cutoffs
```{r}
if (!require("gridExtra")) install.packages("gridExtra")
library(gridExtra)
desired_order = c("<30", "30-74", ">=75")
fit <- survfit(Surv(time_variable, event_variable) ~ factor(pathologist_til_cutoff, levels=desired_order), data = data.all.scores)
group_labels=c("TILs < 30%", "TILs 30% to < 75%", "TILs ≥ 75%")
group_palette=c("darkblue", "darkred", "darkgreen")
filename="km_pathologist_score_standard.png"
plot_custom_km(fit, group_labels, group_palette, filename)
```

KM Plot for pathologist score w/ median cutoffs
```{r}
median_til_score_pathologist = quantile(data.all.scores$gtruth, probs=0.5)
data.all.scores$pathologist_til_cutoff_median <- ifelse(data.all.scores$gtruth <= median_til_score_pathologist, paste0("TILs < ", median_til_score_pathologist), paste0("TILs ≥ ", median_til_score_pathologist))

desired_order = c(paste0("TILs < ", median_til_score_pathologist), paste0("TILs ≥ ", median_til_score_pathologist))
fit <- survfit(Surv(time_variable, event_variable) ~ factor(pathologist_til_cutoff_median, levels=desired_order), data = data.all.scores)

group_labels=desired_order
group_palette=c("darkred", "darkgreen")
filename="km_pathologist_score_median.png"
plot_custom_km(fit, group_labels, group_palette, filename)
```


KM plot with ECTIL score with median cutoff
```{r}
median_ectil_score = round(quantile(data.all.scores[[params$prediction_column]], probs=0.5))

data.all.scores$binary_prediction <- ifelse(data.all.scores[[params$prediction_column]] < median_ectil_score, paste0("ECTILs < ", median_ectil_score), paste0("ECTILs ≥ ", median_ectil_score))

desired_order = c(paste0("ECTILs < ", median_ectil_score), paste0("ECTILs ≥ ", median_ectil_score))

fit <- survfit(Surv(time_variable, event_variable) ~ factor(binary_prediction, levels=desired_order), data = data.all.scores)

group_labels=desired_order
group_palette=c("darkred", "darkgreen")
filename=paste0("km_ectil_", params$prediction_column, "score_median.png")
plot_custom_km(fit, group_labels, group_palette, filename)
```

